#Peer Assignment 1: Analyzing personal activity

This assignment analyzes data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the data. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

##1.Loading and preprocessing the data

The data for this assignment can be downloaded from the github site:

<https://github.com/rdpeng/RepData_PeerAssessment1>

Once you download and unzip dataset *activity.csv* (stored in a comma-separated-value CSV form) in your working directory then load the data using the following command in R. The code below describes how to set your working directory, read the data into the file called "data" and load the necessary libraries that we will need for subsequent analysis.

```{r reading data, results='hide',message=FALSE}
setwd("C:/COURSERA/REPRO_RESEARCH/RepData_PeerAssessment1")

data<-read.csv(unz("C:/COURSERA/REPRO_RESEARCH/RepData_PeerAssessment1/activity.zip","activity.csv"))
library("data.table")
library("dplyr")
library("ggplot2")
library("xtable") # For creating table in markdown
```


The variables included in this dataset are:

* steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
* date: The date on which the measurement was taken in YYYY-MM-DD format
* interval: Identifier for the 5-minute interval in which measurement was taken

There are a total of 17,568 observations in this dataset with some missing.

##2.What is mean total number of steps taken per day?

For this part of the assignment, we ignore the missing values in the dataset.

###2.1.Calculate the total number of steps taken per day

First filter out missing observations and show first few rows of the table

```{r, echo=TRUE}
#Filter NA observations
ccases<-complete.cases(data)
data1<-data[ccases,]
head(data1)
```

Then use dplyr package to calculate total steps per day and see first rows
```{r, echo=TRUE}
perday<-group_by(data1,date) %>% mutate(total_steps=sum(steps))
head(perday)
```


###2.2. Make a histogram of the total number of steps taken each day


```{r histogram, warning=FALSE, message=FALSE}
#Using ggplot
g<-ggplot(perday, aes(total_steps)) 
g+geom_histogram(fill="red")+ 
labs(title = "Total Steps Frequency")+
labs(x="Total Steps")+
labs(y="Count of steps")+
theme_bw(base_family = "Times")
```

###2.3.Calculate and report the mean and median of the total number of steps taken per day

In this section I calculate the mean and median number of steps per day. I thus use the data grouped earlier on per day basis and calculate the mean and median total steps 

```{r calculate mean and median, message=FALSE, warning=FALSE}
mean_total<-mean(perday$total_steps) 
median_total<-median(perday$total_steps) 
mean_total
median_total
```


##3.What is the average daily activity pattern?

###3.1.Time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r, message=FALSE,warning=FALSE}
data1<-as.data.table(data1)
perinterval<-data1[,.(mean_interval = mean(steps)),by=.(interval)]
head(perinterval)
plot(perinterval$interval,perinterval$mean_interval,type="l")
```

###3.2.Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r, message=FALSE, warning=FALSE}
# Find index for max interval
max_int<-max(perinterval$mean_interval)
max_int_index<-perinterval$mean_interval==max_int
perinterval[max_int_index,]
# or get it immediately as follows
perinterval[mean_interval==max(mean_interval),]
```

##4.Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

###4.1.Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r, message=FALSE,warning=FALSE}
missing<-1-sum(complete.cases(data))/nrow(data)
missing
```

###4.2.Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
```{r, message=FALSE,warning=FALSE}
data<-as.data.table(data)
na_index1<-which(is.na(data$steps)) #returns the row index of missing values
na_interval1<-data[na_index1,interval] #saves the interval of each NA missing value
```

###4.3.Create a new dataset that is equal to the original dataset but with the missing data filled in.
``` {r, message=FALSE,warning=FALSE}
test<-data #so that I do not touch original data
test<-as.data.table(test)
class(test$steps)
test1<-test[,.(mean_interval = mean(steps,na.rm=TRUE)),by=.(interval)] #calculate mean steps per intervals and save it in data table
class(test1$mean_interval)
check_interval<-as.integer()
head(test)

test$steps<-as.numeric(test$steps)

for (i in na_index1)
{

  check_interval[i]<-test[i,interval]
  test[i,"steps"]<- test1[interval==check_interval[i],mean_interval]

}
head(test)
```

###4.4.Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r,message=FALSE,warning=FALSE}
perday_new<-group_by(test,date) %>% mutate(total_steps=sum(steps))
head(perday_new)
#Using ggplot to plot per day steps
g<-ggplot(perday_new, aes(total_steps)) 
g+geom_histogram(fill="red")+ 
  labs(title = "Total Steps Frequency")+
  labs(x="Total Steps")+
  labs(y="Count of steps")+
  theme_bw(base_family = "Times")

#Calculate mean and median steps per day
head(perday_new)
mean_total_new<-mean(perday_new$total_steps) 
median_total_new<-median(perday_new$total_steps) 
mean_total_new
median_total_new
```


##5.Are there differences in activity patterns between weekdays and weekends?
For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

###5.1.Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r,message=FALSE,warning=FALSE}
final<-test
final$date<-as.character(final$date)
final$date<-as.Date(final$date,format="%Y-%m-%d")
final$day <- weekdays(as.Date(final$date))

for (i in 1:nrow(final))
{
  if (final$day[i]=="Saturday" | final$day[i]=="Sunday")
  {
    final$weekend_factor[i]="weekend"
  }
  else
  {
    final$weekend_factor[i]="weekday"
  }
  
}
head(final,2000)
```


###5.2.Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r,message=FALSE,warning=FALSE}
final1<-final
final1<-final[,.(avg_steps=mean(steps)),by=.(interval,weekend_factor)]
head(final1)
head(final1)
```

```{r,message=FALSE,warning=FALSE}
gweek<-ggplot(final1,aes(interval,avg_steps))
gweek+geom_point()+
facet_grid(.~weekend_factor)+
geom_line()+
labs(title="Average number of steps per 5 minute interval")+
labs(x="Interval",y="Average Steps")
```
